---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Homework 3A"
author: "Insert Your Name Here"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

library(foreign)
library(epiDisplay)
library(stats)
library(Hmisc)
library(gmodels)

directory <- "/cloud/project/"
```

## Using the dataset “bpa.rda”, answer the following:

# Problem 1. 
## Does urinary BPA differ by age group (20-39, 40-59, 60+)?

```{r problem 1}
#Insert code for problem 1 here
```

# Interpret results from problem 1 here. 

# Problem 2
## Does urinary BPA differ by geneder?

```{r problem 2}
#Insert code for problem 2 here
```
# Interpret results for problem 2 here

# Problem 3
## Does urinary BPA differ by smoking status?

```{r problem 3}
#Insert code for problem 3 here
```
# Interpret results for problem 3 here

# Problem 4
## T2DM can be defined as hemoglobin A1c ≥ 6.5% OR use of diabetes medication. Note: participants with high A1c and taking medication should also be considered T2DM cases.
## First, create a binary variable for T2DM using a1c and dmmed. How many T2DM cases are in each quartile of urinary BPA?

```{r problem 4 variable set up}
#Insert code here to generate T2DM variable and count cases by quartile
```
# Interpret results here


##Compute the odds ratios of T2DM for each of the upper three quartiles of urinary BPA, as compared with the lowest quartile of urinary BPA.
```{r problem 4 calculations}
#Insert code here
```
# Interpret the results for problem 4 here
