---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Hypothesis Testing in R"
author: "Kelly Bakulski, Lauren Middleton"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
---

### Install new packages
```{r install_packages, eval=FALSE, include=FALSE}
# Install packages. Do this only once. 
options(repos="https://cran.rstudio.com" )
install.packages("epiDisplay")
install.packages("here")
install.packages("tidyverse")

# To avoid installing every time: in curly brackets set eval=FALSE
```

### Load packages
```{r load_packages, message=FALSE}

# Load these packages for the current session
library(here)
library(ggcorrplot)
library(gtsummary)
library(epiDisplay)
library(sjlabelled)
library(tidyverse)

```


# Load data
```{r load_data}
# Check the file path
here("nhanes_dataset.rda")
# Load the saved R data
load(here("nhanes_dataset.rda"))
```


# Hypothesis testing in R

## Before you begin: Remember to check your distributions/assumptions
```{r check_distributions}

# Select the continuous variables to check and transform dataset to long form
longer_data <- nhanes_dataset %>%
  dplyr::select(RIDAGEYR,
                INDFMPIR,
                LBDLYMNO,
                LBDNENO,
                LBXRBCSI,
                LBXWBCSI,
                nlr,
                LBXBCD,
                LBXBPB,
                LBXCOT,
                LBXIRN,
                URXUAS) %>%
  remove_all_labels() %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") 

# Plot the distributions of the raw values
longer_data %>%
  ggplot(aes(x = value)) +
    geom_density() +
    facet_wrap(~ variable, scales = "free")
# Are any of these distributions normal?


# Plot the distributions of the log transformed values
longer_data %>%
  ggplot(aes(x = log(value))) +
    geom_density() +
    facet_wrap(~ variable, scales = "free")
# Are any of the log transformed distributions normal?

# Be sure to use your understanding of the distributions when deciding between parametric and non-parametric tests

```


## Correlation tests: Comparing two continuous variables

```{r correlation_tests}

# Correlation test between two continuous variables (default method is Pearson, a parametric test)
cor.test(nhanes_dataset$LBXIRN, nhanes_dataset$LBXRBCSI)

# Can request spearman, a non-parametric test for non-normally distributed variables
cor.test(nhanes_dataset$RIDAGEYR, nhanes_dataset$INDFMPIR, method = "spearman")
```


## Correlation matrix

```{r correlation_matrix}

# Select the columns of interest
nhanes_chems <- nhanes_dataset %>%
  dplyr:: select(LBXIRN,
         LBXCOT,
         LBXBCD,
         LBXBPB,
         LBXWBCSI,
         LBXRBCSI)


# Calculate the correlations, use spearman for non-normally distributed variables
chem_correlations <- cor(nhanes_chems,
                         use = "pairwise.complete.obs",
                         method = "spearman")

# View the correlation matrix
chem_correlations

# Plot the correlations with ggcorrplot
ggcorrplot(chem_correlations,
           type = "lower",
           outline.col = "white",
           lab = TRUE)

```

## Check your understanding
What is the relationship between urinary arsenic (URXUAS) and neutrophil/lymphocyte ratio (nlr)? 
* Make a new code chunk
* What type of variables are they? What are their distributions?
* What is the appropriate test to run?
* Run the test and interpret the output

# Comparing 1 categorical and 1 continuous variable

## T-test: Independent variable dichotomous, dependent variable continuous, parametric test

```{r ttest} 
# The relationship between sex and iron concentration:
# What do you expect?

# Calculate mean of iron concentration by sex
nhanes_dataset %>%
  group_by(sex) %>%
  summarise(mean_iron = mean(LBXIRN, na.rm = TRUE)) %>%
  ungroup()

# T-test of iron and sex: Parametric test with normally distributed dependent variable
t.test(nhanes_dataset$LBXIRN ~ nhanes_dataset$sex) #What do you get? 
# Do they match?

# Calculated within a bivariate table
nhanes_dataset %>%
  dplyr::select(sex,
         LBXIRN) %>%
  tbl_summary(by = sex, #stratify by sex
              statistic = all_continuous() ~ "{mean} ({sd})",
              missing_text = "Missing (n)"
              ) %>%
  add_p(LBXIRN ~ "t.test") %>%
  modify_header(label ~ "**Variable**") %>%
  bold_labels()
```


## Working with a non-normally distributed variable
## T-test and Wilcoxon test: Independent variable dichotomous, dependent variable continuous

```{r ttest_wilcoxon} 

# The relationship between sex and blood Pb
nhanes_dataset %>%
  group_by(sex) %>%
  summarise(mean_lead = mean(LBXBPB, na.rm = TRUE),
            mean_log_lead = mean(log(LBXBPB), na.rm = TRUE)) %>%
  ungroup()

# T-test: Be sure to log transform blood Pb variable before calculating parametric test statistics
t.test(log(nhanes_dataset$LBXBPB) ~ nhanes_dataset$sex)

# Wilcoxon test: Non parametric test, no need to transform variables
wilcox.test(nhanes_dataset$LBXBPB ~ nhanes_dataset$sex)
```


## ANOVA: Independent variable 3+ categories, dependent variable continuous, parametric test

```{r anova}
# The relationship between age and blood Pb:
nhanes_dataset %>%
  group_by(age_groups) %>%
  summarise(mean_lead = mean(LBXBPB, na.rm = TRUE),
            mean_log_lead = mean(log(LBXBPB), na.rm = TRUE)) %>%
  ungroup()

# ANOVA test
aov(log(nhanes_dataset$LBXBPB) ~ nhanes_dataset$age_groups)
anova(aov(log(nhanes_dataset$LBXBPB) ~ nhanes_dataset$age_groups))
# How do we interpret the p-value from an ANOVA test?

# Pairwise t-test
pairwise.t.test(log(nhanes_dataset$LBXBPB), nhanes_dataset$age_groups)

# Pairwise t-test with bonferroni adjustment
pairwise.t.test(log(nhanes_dataset$LBXBPB), nhanes_dataset$age_groups,
                p.adj = "bonferroni")
```


# Comparing two categorical variables

## Chi-square test: Parametric test

```{r chisq_test}
# The relationship between sex and education:
table(nhanes_dataset$sex, nhanes_dataset$education)

# Chi-square test
chisq.test(nhanes_dataset$sex, nhanes_dataset$education)

# Calculated within a bivariate table
nhanes_dataset %>%
  dplyr::select(sex,
         education) %>%
  tbl_summary(by = sex, #stratify by sex
              digits = list(all_categorical() ~ c(0, 1)),
              missing_text = "Missing (n)"
              ) %>%
  add_p() %>%
  modify_header(label ~ "**Variable**") %>%
  bold_labels()

```


##  Fisher's test: Non-parametric test

```{r fisher_test}

# The relationship between sex and iron status:
table(nhanes_dataset$sex, nhanes_dataset$iron_status)

# Fisher exact test
fisher.test(nhanes_dataset$sex, nhanes_dataset$iron_status)

# Calculated within a bivariate table
nhanes_dataset %>%
  dplyr::select(sex,
         iron_status) %>%
  tbl_summary(by = sex, #stratify by sex
              statistic = list(all_categorical() ~ "{n} ({p}%)"), 
              digits = all_categorical() ~ c(0,1),
              missing_text = "Missing (n)"
              ) %>%
  add_p(iron_status ~ "fisher.test") %>%
  modify_header(label ~ "**Variable**") %>%
  bold_labels()

```



# Calculate odds ratios from 2x2 table

```{r odds_ratio}

# Simplify iron into two categories
nhanes_dataset_iron <- nhanes_dataset %>%
  filter(!iron_status == "Excessive") %>%
  droplevels() 
table(nhanes_dataset_iron$iron_status)
levels(nhanes_dataset$iron_status)
levels(nhanes_dataset_iron$iron_status)

nhanes_dataset_iron <- nhanes_dataset_iron %>%
  relevel(iron_status, ref = "Normal")
str(nhanes_dataset_iron$iron_status)
    
# The relationship between iron status (normal and deficient) and sex

# Make 2x2 table
nhanes_dataset_iron %>%
  dplyr::select(sex,
         iron_status) %>%
  tbl_summary(by = sex,
             digits = all_categorical() ~ c(0,1)
              ) %>%
  modify_header(label ~ "**Variable**") %>%
  bold_labels()

# Calculate odds ratio
cc(outcome = nhanes_dataset_iron$iron_status,
   exposure = nhanes_dataset_iron$sex)

```


# Create fake data for graphing odds ratio results

```{r setup_data_for_forestplot}
## Assuming that you obtained ORs and want to compare the results to what was found in other studies.
## Let's create a data frame for these ORs and 95% CIs.

x_num <- c(1, 3, 4, 6, 7)
x1 <- c("Overall", "Study 1", "Study 2", "Study 3", "Study 4")
or <- c(1.5, 1.1, 2.0, 1.4, 1.6)
or_lower_lim <- c(1.2, 0.9, 1.65, 0.95, 1.3)
or_upper_lim <- c(1.85, 1.35, 2.4, 2.05, 2.0)
results <- data.frame(x_num, x1, or, or_lower_lim, or_upper_lim)
results$x1 <- factor(results$x1,
                     levels = results$x1)
```


# Plot odds ratios with a forest plot

```{r forest_plot_ggplot2}
# Forest plot, ggplot2 package
ggplot(results,
       aes(x = or,
           y = x1,
           xmin = or_lower_lim,
           xmax = or_upper_lim)) +
  geom_pointrange() +
  labs(x = "Odds Ratio", y = NULL) +
  geom_vline(xintercept = 1,
             linetype = 2)
```

