---
title: "EPID 674: Data Management in R"
subtitle: "Homework 5 Template"
author: "Kelly Bakulski"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
format: html
editor: visual
---


# Load packages

```{r}
#| label: load_packages
#| echo: true
#| message: false
#| warning: false

# Load packages for this assignment, do this every time
library(here)
library(ggcorrplot)
library(gtsummary)
library(epiDisplay)
library(tidyverse)

```

# Load data
```{r}
#| label: read_data

# Load saved NHANES homework dataset from the project files. This is the dataset we built in homework 2
load(here(("nhanes_homework_dataset.rda")), verbose = TRUE)
```



# 1. Test association between age group and cognition score

### Test the association between age group (variable: age_group) and digit symbol substitution test score (variable: CFDDS).

```{r}
#| label: age_group

# Insert code for question 1 here

```

Interpret the findings: 


# 2. Test association between sex and cognition score

### Test the association between sex group (variable: sex) and digit symbol substitution test score (variable: CFDDS).

```{r}
#| label: sex_group

# Insert code for question 2 here

```

Interpret the findings: 



# 3. Test association between serum cotinine levels and cognition score

### Test the association between serum cotinine (variable: LBXCOT) and digit symbol substitution test score (variable: CFDDS).

```{r}
#| label: cotinine

# Insert code for question 3 here

```

Interpret the findings:  



# 4. Test association between educational attainment and mild cognitive impairment status

### Test the association between educational attainment (variable: education) and mild cognitive impairment status (variable: MCI).

```{r}
#| label: educational_attainment

# Insert code for question 4 here

```

Interpret the findings: 




