---
title: "EPID674 Epidemiologic Data Analysis using R"
subtitle: "Hypothesis Testing in R"
author: "Kelly Bakulski"
date: "Last compiled on `r format(Sys.Date(), '%B %d, %Y')`"
output:
   word_document: default
   html_document: 
    toc: true
    toc_float: true
---

## Chapter 4, Hypothesis Testing with R

# Set up options
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Install new packages
```{r install packages, eval=FALSE}

# Install packages. Do this only once.
options(repos="https://cran.rstudio.com" )
install.packages("epiDisplay")
install.packages("Hmisc")
install.packages("corrplot")

# To avoid installing every time: change set up in curly brackets to eval=FALSE
```

# Load packages
```{r load packages, include=FALSE}
##### Load these packages for the current session
library(epiDisplay)
library(Hmisc)
library(corrplot)
```


# Specify file directories
```{r directories}
directory <- "/cloud/project/" #Class option when coding on RStudio Cloud

# Three alternate options to customize and specify source data directory if programming on a personal computer (only need to do the one that works for you)
# directory<-"M:/EPID674/Data/" # Alternate option 1: location of the parent dataset in afs space of school computer
# directory <- "/Volumes/GoogleDrive/My Drive/Teaching/EPID674/2020_fall/EPID674_Week4_Class/" # Alternate option 2: location of data on my Macbook
# directory<-"C:/Users/bakulski/Documents/EPID674/Data/" #Alternate option 3: location of data on a personal PC

date <- format(Sys.Date(), "%Y%m%d")
```


# Load data, remake useful variables
```{r read data}

# Check the file path
paste0(directory, "nhanes3.rda")
# Load the saved R data
load(paste0(directory, "nhanes3.rda"))
### Or load RData if you saved the workspace image
# load(".RData")


# Remake a few variables from last class if they are no longer in your environment
sex1 <- factor(nhanes$sex, levels = c(1, 2), labels = c("male", "female"))
AGE5b <- cut(nhanes$age, quantile(nhanes$age, c(0, .2, .4, .6, .8, 1)), include.lowest = T) # quintiles
AGE5c <- cut(nhanes$age, breaks = c(19, 40, 50, 60, 70, 90))
age5c <- unclass(AGE5c)
nhanes <- cbind(nhanes, sex1, AGE5b, AGE5c, age5c)
```

# Basic statistical tests in R

# Remember to check your distributions/assumptions
```{r histograms continuous}
hist(nhanes$dbp)
hist(nhanes$bpb)
for (i in 1:ncol(nhanes)) {
  if(is.numeric(nhanes[,i])==T) {
     hist(nhanes[, i], main = colnames(nhanes)[i])
    }
}
```


# Correlation tests
```{r correlation tests}
cor.test(nhanes$age, nhanes$bmi)
cor.test(nhanes$age, nhanes$bpb, method = "spearman")
```

# Correlation matrices
```{r correlation matrix}
nhanes.d <- cbind(nhanes$age, nhanes$bmi, nhanes$bpb, nhanes$sbp)
rcorr(nhanes.d)
rcorr(nhanes.d, type = "spearman")

corrplot(cor(nhanes.d))

nhanes.e <- na.omit(nhanes[, c("age", "bmi", "bpb", "sbp")])
cor(nhanes.e)
nhanes.f <- cbind(nhanes.e$age, nhanes.e$bmi, nhanes.e$bpb, nhanes.e$sbp)
rcorr(nhanes.f)
rcorr(nhanes.f, type = "spearman")

corrplot.mixed(cor(nhanes.f), upper = "ellipse", lower = "number")
```

# T-test and Wilcoxon test
```{r ttest wilcoxon} 
t.test(nhanes$dbp ~ nhanes$sex1)
wilcox.test(nhanes$bpb ~ nhanes$sex1)
t.test(log(nhanes$bpb) ~ nhanes$sex1, var.equal = T)
```

# ANOVA
```{r anova}
tapply(log(nhanes$bpb), nhanes$AGE5b, mean)

aov(log(nhanes$bpb) ~ nhanes$AGE5b)
bpb.aov <- aov(log(nhanes$bpb) ~ nhanes$AGE5b)
anova(bpb.aov)

anova(lm(log(nhanes$bpb) ~ nhanes$AGE5b))
pairwise.t.test(log(nhanes$bpb), nhanes$AGE5b)
pairwise.t.test(log(nhanes$bpb), nhanes$AGE5b, p.adj = "bonf")
```

# Chi-square and Fisher's tests
```{r chisq fisher}
tabpct(nhanes$sex, nhanes$smk)
chisq.test(nhanes$sex, nhanes$smk)

tab1(nhanes$htn)
tabpct(nhanes$race, nhanes$htn)
fisher.test(nhanes$race, nhanes$htn)
```

# Odds Ratios
```{r odds ratio}
cc(nhanes$race, nhanes$htn)
```

# Mantel-Haenszel Odds Ratio
```{r mh odds ratio}
cc(nhanes$htn, nhanes$sex1)
tabpct(nhanes$race, nhanes$sex1)
tabpct(nhanes$htn, nhanes$sex1)

mhor(nhanes$htn, nhanes$race, nhanes$sex1)
```

# Do Exercise 4
